Neural Power Units

### 2020-06-05 papers/main.tex
submitted to nips

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{amsthm}         % definitions
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{caption}
\usepackage{mathtools}

@@

\usepackage{todonotes}

@@

\begin{abstract}
  Conventional Neural Networks can approximate simple arithmetic operations, but fail
  to generalize beyond the range of numbers that were seen during training.
  \emph{Neural Arithmetic Units} aim to overcome this difficulty, but current
  arithmetic units are either limited to operate on positive numbers or can
  only represent a subset of arithmetic operations. We introduce the \emph{Neural
  Power Unit} (NPU) that operates on the
  \emph{full domain of real numbers} and is capable of \emph{learning
  arbitrary power functions} in a single layer.
  The NPU thus fixes the shortcomings of existing arithmetic units and extends
  their expressivity. This is achieved by using formulas from complex arithmetics
  without requiring a conversion of the network to complex numbers.
  We show where the NPU outperforms its
  competitors in terms of accuracy and sparsity on artificial arithmetic
  datasets. Additionally, we demonstrate how the NPU can be used as a highly
  interpretable model to discover the generating equation of a dynamical system
  purely from the data.
\end{abstract}

@@

Numbers and simple algebra are essential not only to human intelligence but
also to the survival of many other
species~\citep{dehaene_number_2011,gallistel_finding_2018}. A successful,
intelligent agent should therefore be able to perform simple arithmetic.  State
of the art neural networks are capable of learning arithmetic, but they
fail to extrapolate beyond the ranges seen during
training~\citep{suzgun_evaluating_2018,lake_generalization_2018}.  The
inability to generalize to unseen inputs is a fundamental problem that hints at
a lack of \emph{understanding} of the given task. The model merely memorizes
the seen inputs and fails to abstract the true learning task.  The failure of
numerical extrapolation on simple arithmetic tasks has been shown
by~\cite{trask_neural_2018}, who also introduce a new class of \emph{Neural
Arithmetic Units} that show good extrapolation performance on some arithmetic
tasks.

Including Neural Arithmetic Units in common neural networks, promises to
significantly increase their extrapolation capabilities due to their inductive
bias towards numerical computation. This is especially true for tasks in which
the data contains mathematical relationships.  Additionally, they promise to
reduce the amount of parameters that are needed for a given task, and can
therefore drastically improve the explainability of the model.  We demonstrate
how neural arithmetic units as part of a \emph{Neural Ordinary Differential
Equation} (NODE, \cite{chen_neural_2019}) can outperform a dense network with
significantly fewer parameters in Sec.~\ref{sub:fractional_sir_model}.  In
fact, our NPU can be used to directly read out the correct generating ODE from
the fitted model.  This is inline with recent efforts to build
\emph{interpretable} models instead of \emph{explaining} black box models
\citep{rudin_stop_2019}, like conventional neural networks.

The currently available arithmetic units all have different strengths and
weaknesses, but none of them solve simple arithmetic completely.
The \emph{Neural Arithmetic Logic Unit} (NALU) by \cite{trask_neural_2018},
chronologically, was the first arithmetic unit. It can solve
addition ($+$, including subtraction), multiplication ($\times$), and division
($\div$), but is limited to positive inputs. The convergence of the
NALU is quite fragile due to an internal gating mechanism between addition and
multiplication paths as well as the use of a logarithm which is problematic for
small inputs.  Recently, \cite{schlor_inalu_2020} introduced the \emph{improved NALU} (iNALU,
to fix the NALU's shortcomings. However, it significantly increases its complexity and we
observe only a slight improvement in performance.\\
\cite{madsen_neural_2020} solve ($+,\times$) with two new units: the
\emph{Neural Addition Unit} (NAU), and the \emph{Neural Multiplication Unit}
(NMU). Instead of gating between addition and multiplication paths, they are
separate units that can be stacked. They can work with the full range real
numbers, converge much more reliably, but are not capable of representing
division.

@@

\textbf{Neural Power Unit.}
We introduce a new arithmetic layer (NPU,
Sec.~\ref{sec:neural_power_unit}) which is capable of learning power
functions $(\prod x_{i}^{w_i})$ for arbitrary real input $x_i$ and power $w_i$ thus including multiplication ($x_1\times x_2 = x_1^1 x_2^1$) as well as division $(x_1\div x_2=x_1^{1}x_2^{-1})$ . This is achieved by using formulas from complex arithmetics (Sec.~\ref{sub:naive_npu}).
Stacks of NAUs and NPUs can thus learn the full spectrum of simple arithmetic operations.

\textbf{Convergence improvement.}
We address the known convergence issues of neural arithmetic units by introducing a
\emph{relevance gate} that smooths out the loss surface of the NPU
(Sec.~\ref{sub:npu}). With the relevance gate the NPU reaches extrapolation and
sparsity errors that are on par with the NMU on ($\times$) and outperform NALU
on ($\div,\sqrt{\cdot}$).

\textbf{Interpretability.}
We show how the NPU can be used as a highly interpretable model for discovery of epidemiological models. 
Specifically, we demontrate its ability to provide a model that can be interpreted as fractional SIR model (Sec.~\ref{sub:fractional_sir_model})
that was used to fit the COVID-19 outbreak in various countries
\citep{taghvaei_fractional_2020}.

@@

A number of different approaches to automatically solve arithmetic tasks have
been studied in the recent years. Approaches include Neural GPUs
\citep{kaiser_neural_2016}, Grid LSTMs \citep{kalchbrenner_grid_2016}, Neural
Turing Machines \citep{graves_neural_2014}, and Neural Random Access Machines
\citep{kurach_neural_2016}. They solve tasks like binary addition and
multiplication, or single digit arithmetic. The Neural Status Register
\citep{faber_neural_2020} focusses on control flow.  The Neural Arithmetic
Expression Calculator \citep{chen_neural_2018}, a hierarchical reinforcement
learner is the only method that solves the division problem, but it operates on
character sequences of arithmetic expressions. Related is symbolic integration
with transformers \citep{lample_deep_2019}. Unfortunately, most of the named models
have severe problems with extrapolation \citep{madsen_measuring_2019,saxton_analysing_2019}.
A solution to the extrapolation problem could be Neural Arithmetic Units. They are
designed with an inductive bias towards systematic, arithmetic computation.
However, currently they are limited in their capabilities of expressing the full
range of simple arithmetic operations ($+,\times,\div$).
In the following two sections we briefly describe the currently available
arithmetic layers including their advantages and drawbacks.

@@

\citet{trask_neural_2018} have demonstrated the severity of the extrapolation
problem of dense networks for even the simplest arithmetic operations, such as
summing or multiplying two numbers.  In order to increase the power of
abstraction to for arithmetic tasks they propose the \emph{Neural Arithmetic
Logic Unit} (NALU) which is capable of learning ($+,\times,\div$).  However,
the NALU cannot handle negative inputs correctly due to the logarithm in
Eq.~\ref{eq:nalu_mult}:
\begin{definition}[{\bf NALU}]
  The NALU consits of a ($+$) and a ($\times$) path that share their weights $\bm{\hat{W}}$ and $\bm{\hat{M}}$.
\begin{align}
  \label{eq:nalu_add}
  \text{Addition: }       & \bm a = \bm W \bm x
                          & \bm{\hat W}& = \tanh(\bm{W}) \odot \sigma(\bm{M}) \\
  \label{eq:nalu_mult}
  \text{Multiplication: } & \bm m = \exp \bm{\hat W}(\log(|\bm x|+\epsilon)) & &\\
  \text{Output: }         & \bm y = \bm a \odot \bm g + \bm m \odot (1-\bm g) 
                          & \bm g& = \sigma(\bm G\bm x)
\end{align}
\end{definition}
Additionally, the logarithm destabilizes training to an extent that the chance of success
can drop below 20\% for ($+,\times$), it becomes practically impossible to
learn ($\div$), and difficult to learn from small inputs in general
\citep{madsen_measuring_2019}.
\cite{schlor_inalu_2020} provide a detailed description of the shortcomings of
the NALU and they suggest an \emph{improved NALU} (iNALU).  The iNALU addresses
the NALU's problems through a number of mechanisms.  It has independent
addition and multiplication weights for Eq.~\ref{eq:nalu_add} and
Eq.~\ref{eq:nalu_mult}, clips weights and gradients to improve training
stability, regularizes the weights to push them away from zero, and, most
importantly, introduces a mechanism to recover the sign that is lost due to the
absolute value in the logarithm. Additionally, the authors propose to
reinitialize the network if its loss is not improving during training.  We
include the iNALU in one of our experiments and find that it only slightly
improves the NALU's performance (Sec.~\ref{sub:simple_arithmetic_task}) at the
cost of a significantly more complicated unit. Our NPU avoids all these mechanisms
by internally using complex arithmetics.

@@

Instead of trying to fix the NALU's convergence issues
\cite{madsen_neural_2020} propose a new unit for ($\times$) only.
The \emph{Neural Multiplication Unit} (NMU) uses explicit multiplications and learns
to gate between identity and ($\times$) of inputs. The NMU is
defined by Eq.~\ref{eq:nmu} and is typically used in conjunction with the
so-called \emph{Neural Addition Unit} (NAU) in Eq.~\ref{eq:nau}.
\begin{definition}[{\bf NMU \& NAU}]
  NMU and NAU are two units that can be stacked to model ($+,\times$).
\begin{align}
  \label{eq:nmu}
  \text{NMU: } &y_j = \prod_i \hat M_{ij} z_{i} + 1 - \hat M_{ij}  & \hat M_{ij}=\min(\max(M_{ij}, 0), 1)\\
  \label{eq:nau}
  \text{NAU: } &\bm y = \bm{\hat{A}} \bm x & \hat A_{ij}=\min(\max(A_{ij}, -1), 1)
\end{align}
\end{definition}
Both NMU and NAU are regularized with $\mathcal{R} = \sum_{ij} \min(|W_{ij}|,
|1-W_{ij}|)$ and their weights are clipped, which biases them towards learning
an operation or pruning it completely.  The combination of NAU and NMU can thus
learn $(+,\times)$ for both positive and negative inputs. Training NAU and NMU
is stable and succeeds much more frequently than with the NALU, but they cannot
represent ($\div$), which we address with our NPU.

@@

To fix the deficiencies of current arithmetic units, we propose a new
arithmetic unit (inspired by NALU) that can learn arbitrary power functions $(x^w)$ (including
$\times,\div$) for positive and negative numbers, and still train well.
Combined with the NAU we solve the full range of arithmetic operations.
This is possible through a simple modification of the ($\times$)-path of the
NALU (Eq.~\ref{eq:nalu_mult_2}).  We suggest to use the complex logarithm and
to allow $\bm W$ to become complex.  The complex logarithm is also defined for
negative inputs and a complex $\bm W$ aides convergence (see
Sec.~\ref{sub:fractional_sir_model}). The improvement during training might be
explained by the additional imaginary parameters that make it possible to avoid
regions with an uninformative gradient signal.

@@

With the modifications introduced above we can extend the multiplication path
of the NALU from
\begin{align}
  \label{eq:nalu_mult_2}
  \bm m = \exp \bm W(\log_{\text{real}}(|\bm x|+\epsilon))
\end{align}
to use the complex logarithm ($\log\coloneqq\log_{\text{complex}}$) and a
complex weight $\bm W$ to
\begin{align}
  \label{eq:1}
  \bm z = \exp(\bm W\log \bm x) = \exp\left((\bm \Wre + i\Wim) \log\bm x\right).
\end{align}
The complex log in Eq.~\ref{eq:1} lifts the positivity constraint on $\bm x$ resulting
in a layer that can process both positive and negative numbers correctly.  A complex
weight matrix $\bm W$ somewhere in a larger network would result in complex
gradients in other layers.  This would effectively result in doubling the
number of parameters of the whole network. As we are only interested in real
networks outputs, we can avoid this doubling by considering only the real part
of the output $\bm z$:
\begin{align}
  \real(\bm z) &= \real(\exp((\Wre + i\Wim)(\log\bm r + i\pi\bm k))) \\
    \label{eq:npumult}
    &= \exp(\Wre\log\bm r - \pi\Wim\bm k) \odot \cos(\Wim\log\bm r + \pi\Wre\bm k).
\end{align}
Above we have used Euler's formula\footnote{Euler's formula states that for any
real number $x$: $e^{ix} = \cos x + i\sin x$} and the fact that the complex
logarithm for real valued inputs is
\begin{align}
  \log x = \log r + i\theta = \log r + ik\pi,
\end{align}
where $k=0$ if $r\geq0$ and $k=1$ if $r<0$.
A diagram of the NaiveNPU is shown in Fig.~\ref{fig:npu_diagram}.
\begin{definition}[{\bf NaiveNPU}]
  The Naive Neural Power Unit with matrices $\Wre$ and
  $\Wim$ representing real and imaginary part of the complex numbers defined as
  \begin{gather}
    \label{eq:naivenpu_def}
    \bm z = \exp(\Wre \log\bm r - \Wim\bm k) \odot \cos(\Wim\log \bm r + \Wre\bm k), \text{ where }\\
    \nonumber
    \bm r = |\bm x|,
    \quad
    k_i = \begin{cases}
       0  & x_i \leq 0 \\
       \pi & x_i > 0
    \end{cases}
  \end{gather}
\end{definition}

\begin{figure}
  \begin{minipage}{.45\textwidth}
      \resizebox{0.9\textwidth}{!}{\input{../tikz/npu-diagram.tex}}
      \captionof{figure}{NaiveNPU diagram, with input $\bm x$ and output $\bm y$.
      Vectors in green, trainables in orange, functions in blue.}
      \label{fig:npu_diagram}
  \end{minipage}
  \hspace{0.01\textwidth}
  \begin{minipage}{0.54\textwidth}
    \resizebox{\textwidth}{!}{\input{../tikz/gatednpu-diagram.tex}}
    \captionof{figure}{NPU diagram. The NPU has a relevance gate $\bm g$
    (hatched background) in front of the input to the actual unit to prevent
    vanishing gradients.}
    \label{fig:gatednpu_diagram}
  \end{minipage}
\end{figure}

@@

The NaiveNPU has difficulties to converge on large scale tasks, and to reach
sparse results in cases where the input to a given row is small. We
demonstrate this on a toy example of learning the identity on one of two inputs
(neglecting the second one):
\begin{align}
  \nonumber
  \mathcal{L} &= |\text{m}(x_1,x_2) - x_1|, \quad \text{where}
  & \text{m} = \npu(2,1), \quad
  &x_1 \sim \mathcal U(0,2), \quad x_2 \sim \mathcal U(0,0.05).
\end{align}
The left plot in Fig.~\ref{fig:id_loss} depicts
the the norm of gradient of a batch of two dimensional inputs, one of which is small and
irrelevant. It is clearly visible that even for this simple example, a large
part of the gradient surface of the NaiveNPU is zero.  This vanishing gradient can be explained
as follows. One row of NaiveNPU weights effectively raises each input to a power and
multiplies them: $x_1^{w_1} x_2^{w_2} \dots x_n^{w_n}$. If a single input $x_i$
is constantly close to zero (i.e. irrelevant), the whole row will be zero, no
matter what its weights are and the gradient information on all other weights
is lost.  We therefore introduce a gate on the input of the NPU that can turn
irrelevant inputs into 1s.
A diagram of the NPU is shown in Fig.~\ref{fig:gatednpu_diagram}.
\begin{definition}[{\bf NPU}]
  The NPU extends the NaiveNPU by the relevance gate $\bm g$ on the input $\bm x$.
   \begin{gather}
    \label{eq:gatednpu_def}
    \bm z = \exp(\Wre \log\bm r - \Wim\bm k) 
          \odot \cos(\Wim\log \bm r + \Wre\bm k), \text{ where } \\
    \bm r = \bm{\hat g} \odot |\bm x| + (1-\bm{\hat g}),
    \quad
    k_i = \begin{cases}
       0  & x_i \leq 0 \\
      \hat g_i\pi & x_i > 0
    \end{cases},
    \quad
    \hat g_i = \min(\max(g_i,0),1)
  \end{gather}
\end{definition}
\begin{figure}
  \centering
  \begin{minipage}{0.55\textwidth}
    \includegraphics[width=\textwidth]{npu_gatednpu_id_grad.pdf}
  \end{minipage}
  \begin{minipage}{0.42\textwidth}
  \end{minipage}
  \caption{Absolute value of a gradient of NaiveNPU and NPU for the task
  of learning the identity on $x_1$. Inputs and
  loss are defined on the right, gradient surfaces on the left
  (black areas are beyond the color scale). The correct solution is
  $w_1=1$ and $w_2=0$.  The NaiveNPU has a large zero gradient region for
  $w_2>0.75$, while the NPU's surface is much more informative. The gates
  for central plot are fixed at $g_1=g_2=0.5$ which corresponds to the initial
  gate parameters. During training they will adjust as needed, in this case to
  $g_1=1$ and $g_2=0$. $\bm W_i$ is set to zero in all plots.
  }%
  \label{fig:id_loss}
\end{figure}
The central plot of Fig.~\ref{fig:id_loss} shows the gradient of the NPU
on the identity task with its initial gate setting of $g_1=g_2=0.5$. The
large zero-gradient region of the NaiveNPU is gone. The last plot shows the
same loss for $g_1=1$ and $g_2=0$, which corresponds to the correct gates at
the end of NPU training. The gradient is independent of $w_2$, which
means that it can easily be pruned by a simple regularization such as L1.  In
Sec.~\ref{sub:large_scale_arithmetic_task} we show how important the relevance
gating mechanism is for the convergence and sparsity of large models.  Sparsity
is especially important in order to use the NPU as an interpretable model.

>>>>

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{amsthm}         % definitions
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{caption}
\usepackage{mathtools} 
\usepackage{amsthm}

\newtheorem*{remark}{Remark}

@@

@@

\begin{abstract}
  Conventional Neural Networks can approximate simple arithmetic operations,
  but fail to generalize beyond the range of numbers that were seen during
  training.  \emph{Neural Arithmetic Units} aim to overcome this difficulty,
  but current arithmetic units are either limited to operate on positive
  numbers or can only represent a subset of arithmetic operations. We introduce
  the \emph{Neural Power Unit} (NPU) that operates on the \emph{full domain of
  real numbers} and is capable of \emph{learning arbitrary power functions} in
  a single layer.  The NPU thus fixes the shortcomings of existing arithmetic
  units and extends their expressivity. We achieve this by using
  complex arithmetic without requiring a conversion of the network to complex
  numbers. A simplification of the unit to the \emph{RealNPU} yields a highly interpretable model.
  We show that the NPUs outperform their competitors in terms of
  accuracy and sparsity on artificial arithmetic datasets, and that the RealNPU can discover
  the governing equations of a dynamical systems only from data.
\end{abstract}

@@

Numbers and simple algebra are essential not only to human intelligence but
also to the survival of many other
species~\citep{dehaene_number_2011,gallistel_finding_2018}. A successful,
intelligent agent should, therefore, be able to perform simple arithmetic.  State
of the art neural networks are capable of learning arithmetic, but they
fail to extrapolate beyond the ranges seen during
training~\citep{suzgun_evaluating_2018,lake_generalization_2018}.  The
inability to generalize to unseen inputs is a fundamental problem that hints at
a lack of \emph{understanding} of the given task. The model merely memorizes
the seen inputs and fails to abstract the true learning task.  The failure of
numerical extrapolation on simple arithmetic tasks has been shown
by~\cite{trask_neural_2018}, who also introduced a new class of \emph{Neural
Arithmetic Units} with good extrapolation performance on some arithmetic
tasks.

Including Neural Arithmetic Units in standard neural networks, promises to
significantly increase their extrapolation capabilities due to their inductive
bias towards numerical computation. This is especially true for tasks in which
the data generating process contains mathematical relationships.  Additionally,
they promise to reduce the number of parameters that are needed for a given
task, and can therefore drastically improve the explainability of the model.
We demonstrate how a \emph{Neural Ordinary Differential Equation} (NODE,
\cite{chen_neural_2019}) composed of small neural arithmetic units can
outperform a much larger NODE built from dense layers (Sec.~\ref{sub:fractional_sir_model}).
In fact, our new unit can be used to directly read out the correct generating ODE
from the fitted model.  This is in line with recent efforts to build
\emph{interpretable} models instead of \emph{explaining} black-box models
\citep{rudin_stop_2019}, like conventional neural networks.

The currently available arithmetic units all have different strengths and
weaknesses, but none of them solve simple arithmetic completely.
The \emph{Neural Arithmetic Logic Unit} (NALU) by \cite{trask_neural_2018},
chronologically, was the first arithmetic unit. It can solve
addition ($+$, including subtraction), multiplication ($\times$), and division
($\div$), but is limited to positive inputs. The convergence of the
NALU is quite fragile due to an internal gating mechanism between addition and
multiplication paths as well as the use of a logarithm which is problematic for
small inputs.  Recently, \cite{schlor_inalu_2020} introduced the \emph{improved NALU} (iNALU,
to fix the NALU's shortcomings. However, it significantly increases its complexity, and we
observe only a slight improvement in performance.
\cite{madsen_neural_2020} solve ($+,\times$) with two new units: the
\emph{Neural Addition Unit} (NAU), and the \emph{Neural Multiplication Unit}
(NMU). Instead of gating between addition and multiplication paths, they are
separate units that can be stacked. They can work with the full range of real
numbers, converge much more reliably, but cannot represent
division.

@@

\textbf{Neural Power Unit.}
We introduce a new arithmetic layer (NPU, Sec.~\ref{sec:neural_power_unit})
which is capable of learning products of power functions $(\prod x_{i}^{w_i})$ of
arbitrary real inputs $x_i$ and power $w_i$, thus including multiplication
($x_1\times x_2 = x_1^1 x_2^1$) as well as division $(x_1\div
x_2=x_1^{1}x_2^{-1})$. This is achieved by using formulas from complex
arithmetic (Sec.~\ref{sub:naive_npu}).  Stacks of NAUs and NPUs can thus learn
the full spectrum of simple arithmetic operations.

\textbf{Convergence improvement.}
We address the known convergence issues of neural arithmetic units by
introducing a \emph{relevance gate} that smooths out the loss surface of the
NPU (Sec.~\ref{sub:npu}). With the relevance gate, which helps to learn to
ignore variables, the NPU reaches extrapolation error and sparsity that is on
par with the NMU on ($\times$) and outperform NALU on ($\div,\sqrt{\cdot}$).

\textbf{Interpretability.}
We show how a power unit can be used as a highly interpretable model for equation discovery
of dynamical systems. Specifically, we demonstrate its ability to identify
a model that can be interpreted as a SIR model with fractional powers
(Sec.~\ref{sub:fractional_sir_model}) that was used to fit the COVID-19
outbreak in various countries \citep{taghvaei_fractional_2020}.

@@

Several different approaches to automatically solve arithmetic tasks have
been studied in recent years. Approaches include Neural GPUs
\citep{kaiser_neural_2016}, Grid LSTMs \citep{kalchbrenner_grid_2016}, Neural
Turing Machines \citep{graves_neural_2014}, and Neural Random Access Machines
\citep{kurach_neural_2016}. They solve tasks like binary addition and
multiplication, or single-digit arithmetic. The Neural Status Register
\citep{faber_neural_2020} focusses on control flow.  The Neural Arithmetic
Expression Calculator \citep{chen_neural_2018}, a hierarchical reinforcement
learner, is the only method that solves the division problem, but it operates on
character sequences of arithmetic expressions. Related is symbolic integration
with transformers \citep{lample_deep_2019}. Unfortunately, most of the named models
have severe problems with extrapolation \citep{madsen_measuring_2019,saxton_analysing_2019}.
A solution to the extrapolation problem could be Neural Arithmetic Units. They are
designed with an inductive bias towards systematic, arithmetic computation.
However, currently, they are limited in their capabilities of expressing the full
range of simple arithmetic operations ($+,\times,\div$).
In the following two sections, we briefly describe the currently available
arithmetic layers, including their advantages and drawbacks.

@@

\citet{trask_neural_2018} have demonstrated the severity of the extrapolation
problem of dense networks for even the simplest arithmetic operations, such as
summing or multiplying two numbers.  In order to increase the power of
abstraction for arithmetic tasks, they propose the \emph{Neural Arithmetic
Logic Unit} (NALU), which is capable of learning ($+,\times,\div$).  However,
the NALU cannot handle negative inputs correctly due to the logarithm in
Eq.~\ref{eq:nalu_mult}:
\begin{definition}[{\bf NALU}]
  The NALU consits of a ($+$) and a ($\times$) path that share their weights $\bm{\hat{W}}$ and $\bm{\hat{M}}$.
\begin{align}
  \label{eq:nalu_add}
  \text{Addition: }       & \bm a = \bm W \bm x
                          & \bm{\hat W}& = \tanh(\bm{W}) \odot \sigma(\bm{M}) \\
  \label{eq:nalu_mult}
  \text{Multiplication: } & \bm m = \exp \bm{\hat W}(\log(|\bm x|+\epsilon)) & &\\
  \text{Output: }         & \bm y = \bm a \odot \bm g + \bm m \odot (1-\bm g) 
                          & \bm g& = \sigma(\bm G\bm x)
\end{align}
\end{definition}
Additionally, the logarithm destabilizes training to the extent that the chance of success
can drop below 20\% for ($+,\times$), it becomes practically impossible to
learn ($\div$) and difficult to learn from small inputs in general
\citep{madsen_measuring_2019}.
\cite{schlor_inalu_2020} provide a detailed description of the shortcomings of
the NALU, and they suggest an \emph{improved NALU} (iNALU).  The iNALU addresses
the NALU's problems through several mechanisms.  It has independent
addition and multiplication weights for Eq.~\ref{eq:nalu_add} and
Eq.~\ref{eq:nalu_mult}, clips weights and gradients to improve training
stability, regularizes the weights to push them away from zero, and, most
importantly, introduces a mechanism to recover the sign that is lost due to the
absolute value in the logarithm. Additionally, the authors propose to
reinitialize the network if its loss is not improving during training.  We
include the iNALU in one of our experiments and find that it only slightly
improves the NALU's performance (Sec.~\ref{sub:simple_arithmetic_task}) at the
cost of a significantly more complicated unit. Our NPU avoids all these mechanisms
by internally using complex arithmetics.

@@

Instead of trying to fix the NALU's convergence issues,
\cite{madsen_neural_2020} propose a new unit for ($\times$) only.
The \emph{Neural Multiplication Unit} (NMU) uses explicit multiplications and learns
to gate between identity and ($\times$) of inputs. The NMU is
defined by Eq.~\ref{eq:nmu} and is typically used in conjunction with the
so-called \emph{Neural Addition Unit} (NAU) in Eq.~\ref{eq:nau}.
\begin{definition}[{\bf NMU \& NAU}]
  NMU and NAU are two units that can be stacked to model ($+,\times$).
\begin{align}
  \label{eq:nmu}
  \text{NMU: } &y_j = \prod_i \hat M_{ij} z_{i} + 1 - \hat M_{ij}  & \hat M_{ij}=\min(\max(M_{ij}, 0), 1)\\
  \label{eq:nau}
  \text{NAU: } &\bm y = \bm{\hat{A}} \bm x & \hat A_{ij}=\min(\max(A_{ij}, -1), 1)
\end{align}
\end{definition}
Both NMU and NAU are regularized with $\mathcal{R} = \sum_{ij} \min(|W_{ij}|,
|1-W_{ij}|)$, and their weights are clipped, which biases them towards learning
an operation or pruning it completely.  The combination of NAU and NMU can thus
learn $(+,\times)$ for both positive and negative inputs. Training NAU and NMU
is stable and succeeds much more frequently than with the NALU, but they cannot
represent ($\div$), which we address with our NPU.

@@

To fix the deficiencies of current arithmetic units, we propose a new
arithmetic unit (inspired by NALU) that can learn arbitrary products of power functions $(\prod x_{i}^{w_i})$ (including
$\times,\div$) for positive and negative numbers, and still train well.
Combined with the NAU, we solve the full range of arithmetic operations.
This is possible through a simple modification of the ($\times$)-path of the
NALU (Eq.~\ref{eq:nalu_mult_2}). 
We suggest to replace the logarithm of the absolute value by the complex logarithm and
to allow $\bm W$ to be complex as well. Since the complex logarithm is defined for
negative inputs, the NPU does not have a problem with negative numbers. A complex $\bm W$ 
improves convergence at the expense of explainability (see
Sec.~\ref{sub:fractional_sir_model}). The improvement during training might be
explained by the additional imaginary parameters that make it possible to avoid
regions with an uninformative gradient signal.

@@

With the modifications introduced above we can extend the multiplication path
of the NALU from
\begin{align}
  \label{eq:nalu_mult_2}
  \bm m = \exp \bm W(\log_{\text{real}}(|\bm x|+\epsilon))
\end{align}
to use the complex logarithm ($\log\coloneqq\log_{\text{complex}}$) and a
complex weight $\bm W$ to
\begin{align}
  \label{eq:1}
  \bm z = \exp(\bm W\log \bm x) = \exp\left((\bm \Wre + i\Wim) \log\bm x\right).
\end{align}
The complex log in Eq.~\ref{eq:1} lifts the positivity constraint on $\bm x$ resulting
in a layer that can process both positive and negative numbers correctly.  A complex
weight matrix $\bm W$ somewhere in a larger network would result in complex
gradients in other layers.  This would effectively result in doubling the
number of parameters of the whole network. As we are only interested in real
networks outputs, we can avoid this doubling by considering only the real part
of the output $\bm z$:
\begin{align}
  \real(\bm z) &= \real(\exp((\Wre + i\Wim)(\log\bm r + i\pi\bm k))) \\
    \label{eq:npumult}
    &= \exp(\Wre\log\bm r - \pi\Wim\bm k) \odot \cos(\Wim\log\bm r + \pi\Wre\bm k).
\end{align}
Above we have used Euler's formula\footnote{Euler's formula states that for any
real number $x$: $e^{ix} = \cos x + i\sin x$} and the fact that the complex
logarithm for real valued inputs is
\begin{align}
  \log x = \log r + i\theta = \log r + ik\pi,
\end{align}
where $k=0$ if $r\geq0$ and $k=1$ if $r<0$.
A diagram of the NaiveNPU is shown in Fig.~\ref{fig:npu_diagram}.
\begin{definition}[{\bf NaiveNPU}]
  The Naive Neural Power Unit with matrices $\Wre$ and
  $\Wim$ representing real and imaginary part of the complex numbers defined as
  \begin{gather}
    \label{eq:naivenpu_def}
    \bm z = \exp(\Wre \log\bm r - \pi\Wim\bm k) \odot \cos(\Wim\log \bm r + \pi\Wre\bm k), \text{ where }\\
    \nonumber
    \bm r = |\bm x|,
    \quad
    k_i = \begin{cases}
       0 & x_i \leq 0 \\
       1 & x_i > 0
    \end{cases}
  \end{gather}
\end{definition}

\begin{figure}
  \begin{minipage}{.45\textwidth}
      \resizebox{0.9\textwidth}{!}{\input{../tikz/npu-diagram.tex}}
      \captionof{figure}{NaiveNPU diagram, with input $\bm x$ and output $\bm y$.
      Vectors in green, trainables in orange, functions in blue.}
      \label{fig:npu_diagram}
  \end{minipage}
  \hspace{0.01\textwidth}
  \begin{minipage}{0.54\textwidth}
    \resizebox{\textwidth}{!}{\input{../tikz/gatednpu-diagram.tex}}
    \captionof{figure}{NPU diagram. The NPU has a relevance gate $\bm g$
    (hatched background) in front of the input to the actual unit to prevent
    zero gradients.}
    \label{fig:gatednpu_diagram}
  \end{minipage}
\end{figure}

@@

The NaiveNPU has difficulties to converge on large scale tasks, and to reach
sparse results in cases where the input to a given row is small. We
demonstrate this on a toy example of learning the identity on one of two inputs
(neglecting the second one):
\begin{align}
  \nonumber
  \mathcal{L} &= |\text{m}(x_1,x_2) - x_1|, \quad \text{where}
  & \text{m} = \npu(2,1), \quad
  &x_1 \sim \mathcal U(0,2), \quad x_2 \sim \mathcal U(0,0.05).
\end{align}
The left plot in Fig.~\ref{fig:id_loss} depicts the gradient norm of NPU and
NaiveNPU for a batch of two-dimensional inputs. One input is small and
irrelevant. Even in this simple example, gradient of the NaiveNPU is close to
zero in large parts of the parameter space This can be explained as follows.
One row of NaiveNPU weights effectively raises each input to a power and
multiplies them: $x_1^{w_1} x_2^{w_2} \dots x_n^{w_n}$. If a single input $x_i$
is constantly close to zero (i.e. irrelevant), the whole row will be zero, no
matter what its weights are and the gradient information on all other weights
is lost.  Therefore, we introduce a gate on the input of the NPU that can turn
irrelevant inputs into 1s.
A diagram of the NPU is shown in Fig.~\ref{fig:gatednpu_diagram}.
\begin{definition}[{\bf NPU}]
  The NPU extends the NaiveNPU by the relevance gate $\bm g$ on the input $\bm x$.
   \begin{gather}
    \label{eq:gatednpu_def}
    \bm z = \exp(\Wre \log\bm r - \pi\Wim\bm k) 
          \odot \cos(\Wim\log \bm r + \pi\Wre\bm k), \text{ where } \\
    \bm r = \bm{\hat g} \odot |\bm x| + (1-\bm{\hat g}),
    \quad
    k_i = \begin{cases}
       0  & x_i \leq 0 \\
      \hat g_i & x_i > 0
    \end{cases},
    \quad
    \hat g_i = \min(\max(g_i,0),1)
  \end{gather}
\end{definition}
\begin{figure}
  \centering
  \begin{minipage}{0.55\textwidth}
    \includegraphics[width=\textwidth]{npu_gatednpu_id_grad.pdf}
  \end{minipage}
  \begin{minipage}{0.42\textwidth}
  \end{minipage}
  \caption{Norm of the gradient of NaiveNPU and NPU for the task
  of learning the identity on $x_1$. Inputs and
  loss are defined on the right, gradient surfaces on the left
  (black areas are beyond the color scale). The correct solution is
  $w_1=1$ and $w_2=0$.  The NaiveNPU has a large zero gradient region for
  $w_2>0.75$, while the NPU's surface is much more informative. The gates
  for central plot are fixed at $g_1=g_2=0.5$ which corresponds to the initial
  gate parameters. During training they will adjust as needed, in this case to
  $g_1=1$ and $g_2=0$. $\bm W_i$ is set to zero in all plots.
  }%
  \label{fig:id_loss}
\end{figure}
The central plot of Fig.~\ref{fig:id_loss} shows the gradient of the NPU
on the identity task with its initial gate setting of $g_1=g_2=0.5$. The
large zero-gradient region of the NaiveNPU is gone. The last plot shows the
same loss for $g_1=1$ and $g_2=0$, which corresponds to the correct gates at
the end of NPU training. The gradient is independent of $w_2$, which
means that it can easily be pruned by a simple regularization such as L1.  In
Sec.~\ref{sub:large_scale_arithmetic_task} we show how important the relevance
gating mechanism is for the convergence and sparsity of large models.  Sparsity
is especially important in order to use the NPU as an interpretable model.

\textbf{Initialization}%
\label{sub:initialization}
We recommend initializing the NPU with a Glorot Uniform distribution on the
real weights $\Wre$. The imaginary weights $\Wim$ can be initialized to zeros,
so they will only be used where necessary, and the gate $\bm g$ with 0.5, so
the NPU can choose to output 1.
